Mergeable Types for Highly-Available Replicated Data

### 2017-02-26 paper/dali.tex
List eg.

% Formatting commands
% -------------------
\newcommand{\code}[1]{\,{\tt #1}\,}
\newcommand{\spc}[0]{\quad}
\newcommand{\ALT}{~\mid~}
\newcommand{\rel}[1]{{R}_{\mathit{#1}}}
\newcommand{\conj}{~\wedge~}
\newcommand{\disj}{~\vee~}
\newcommand{\rulelabel}[1]{\textrm{\sc {#1}}}
\newcommand{\ilrulelabel}[1]{{\sc #1}}
\newcommand{\RULE}[2]{\frac{\begin{array}{c}#1\end{array}}
                           {\begin{array}{c}#2\end{array}}}
\newcommand{\denot}[1]{\llbracket #1 \rrbracket}
\newcommand{\bind}{>\!\!>\!\!=}
\newcommand{\coloneqq}{::=}
\newcommand{\stepsto}{\longrightarrow}
\newcommand{\return}[1]{\C{return}\;#1}
\newcommand{\run}[2]{\C{run}\;#1\;#2}
\newcommand{\fork}[1]{\C{fork}\;#1}
\newcommand{\push}[1]{\C{push}\;#1}
\newcommand{\pull}{\C{pull}}
\newcommand{\semsucceq}{\succeq_{\circ}}
\newcommand{\sempreceq}{\preceq_{\circ}}
\newcommand{\under}[2]{#1\,\vdash\,#2}
\newcommand{\mbleto}{\rightsquigarrow_m}

% New colors
%------------
\definecolor{Bittersweet}{rgb}{1.0, 0.44, 0.37}

% Listings Code
%---------------
\newcommand{\lstml}{
\lstset{ %
language=ML, % choose the language of the code
basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
keywordstyle=\color{Bittersweet},
% numbers=left,                   % where to put the line-numbers
numberstyle=\tiny,      % the size of the fonts that are used for the line-numbers
stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
numbersep=5pt,                  % how far the line-numbers are from the code
showspaces=false,               % show spaces adding particular underscores
showstringspaces=false,         % underline spaces within strings
showtabs=false,                 % show tabs within strings adding particular underscores
% frame=single,                   % adds a frame around the code
tabsize=2,                      % sets default tabsize to 2 spaces
captionpos=b,                   % sets the caption-position to bottom
breaklines=true,                % sets automatic line breaking
breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
commentstyle=\itshape\color{MidnightBlue},
%escapeinside={\%*}{*)},         % if you want to add a comment within your code
morekeywords={module, @@deriving, not, : , /\\}
}}
\lstnewenvironment{ocaml}
    { % \centering
			\lstml
      \lstset{}%
      \csname lst@setfirstlabel\endcsname}
    { %\centering
      \csname lst@savefirstlabel\endcsname}

@@

%% Title information
\title[DaLi]{DaLi: Database as a Library}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

>>>>

% Formatting commands
% -------------------
\newcommand{\code}[1]{\,{\tt #1}\,}
\newcommand{\spc}[0]{\quad}
\newcommand{\ALT}{~\mid~}
\newcommand{\rel}[1]{{R}_{\mathit{#1}}}
\newcommand{\conj}{~\wedge~}
\newcommand{\disj}{~\vee~}
\newcommand{\rulelabel}[1]{\textrm{\sc {#1}}}
\newcommand{\ilrulelabel}[1]{{\sc #1}}
\newcommand{\RULE}[2]{\frac{\begin{array}{c}#1\end{array}}
                           {\begin{array}{c}#2\end{array}}}
\newcommand{\denot}[1]{\llbracket #1 \rrbracket}
\newcommand{\bind}{>\!\!>\!\!=}
\newcommand{\coloneqq}{::=}
\newcommand{\stepsto}{\longrightarrow}
\newcommand{\return}[1]{\C{return}\;#1}
\newcommand{\run}[2]{\C{run}\;#1\;#2}
\newcommand{\fork}[1]{\C{fork}\;#1}
\newcommand{\push}[1]{\C{push}\;#1}
\newcommand{\pull}{\C{pull}}
\newcommand{\semsucceq}{\succeq_{\circ}}
\newcommand{\sempreceq}{\preceq_{\circ}}
\newcommand{\under}[2]{#1\,\vdash\,#2}
\newcommand{\mbleto}{\rightsquigarrow_m}
\newcommand{\drawsome}{{\sf Canvas}\xspace}
\newcommand{\lang}{\lambda_b}
\newcommand{\length}[1]{\left|#1\right|}

% New colors
%------------
\definecolor{Bittersweet}{rgb}{1.0, 0.44, 0.37}
\definecolor{MidnightBlue}{rgb}{0.0, 0.2, 0.4}

% Listings Code
%---------------
\newcommand{\lstml}{
\lstset{ %
language=ML, % choose the language of the code
basicstyle=\footnotesize\ttfamily,       % the size of the fonts that are used for the code
keywordstyle=\color{Bittersweet},
% numbers=left,                   % where to put the line-numbers
numberstyle=\tiny,      % the size of the fonts that are used for the line-numbers
stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
numbersep=5pt,                  % how far the line-numbers are from the code
showspaces=false,               % show spaces adding particular underscores
showstringspaces=false,         % underline spaces within strings
showtabs=false,                 % show tabs within strings adding particular underscores
% frame=single,                   % adds a frame around the code
tabsize=2,                      % sets default tabsize to 2 spaces
captionpos=b,                   % sets the caption-position to bottom
breaklines=true,                % sets automatic line breaking
breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
commentstyle=\itshape\color{MidnightBlue},
%escapeinside={\%*}{*)},         % if you want to add a comment within your code
morekeywords={module, when, @@deriving, not, : , /\\}
}}
\lstnewenvironment{ocaml}
    { % \centering
			\lstml
      \lstset{}%
      \csname lst@setfirstlabel\endcsname}
    { %\centering
      \csname lst@savefirstlabel\endcsname}

@@

%% Title information
\title[]{Mergeable Types for Highly-Available Replicated Data}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

### 2017-02-26 paper/mergeable-types.tex
List eg.

We now formally define what we mean by a mergeable type. The
definition serves as a guide to write merge functions for functional
data types, elevating them to the level of replicated data types.

A functional data type library is a tuple consisting of a type ($t$),
and a collection of functions ($f$, $g$, etc.) of type $t \rightarrow
t$ that are \emph{primitive morphisms}. A morphism ($P$, $Q$, etc.) is
either a primitive morphism, or an associative composition of
morphisms ($P \circ Q \circ R$). An object $B:t$ is called a
\emph{semantic successor} of $A:t$ (conversely, $A$ is called a
\emph{semantic ancestor} of $B$) if and only if there exists a
morphism $P$ such that $P(A) = B$.

The aim of a three-way merge function over a type $t$ is to merge a
pair of semantic successors, $B:t$ and $C:t$, of an object $A:t$, into
another object $D:t$ such that the relationship between the semantic
successors and $D$ satisfies certain conditions. These conditions can
be understood by observing the \rulelabel{E-Pull-Wait} rule of
Fig.~\ref{fig:opsem}, which applies the \C{merge} function to a pair
of concurrent versions ($v_1$ and $v_2$) and their least common
ancestor ($v$). Thus, the only relationship that exists between $v$
and $v_1$, and $v$ and $v_2$ is the syntactic ancestor relationship
that follows from the computation's branching structure. The merge
function, as described above, however assumes that concurrent versions
are semantic successors of their LCA.  It is therefore essential to
maintain coherence between the syntactic and semantic ancestor
relations.

The ways in which syntactic ancestor relationships are created among
versions is captured in Fig.~\ref{fig:syntactic-ancestors}. Whenever an
object is pushed onto the branch, an ancestor relationship is created
between the previous version $v_1$, and the newly created version
$v_2$ (Fig.~\ref{fig:syntactic-ancestor-1}). However, since $v_2$ is
pushed by the thread after reading $v_1$ into the heap, it is
reasonable to assume that $v_2$ is a result of applying a morphism $P$
to $v_1$ (i.e., $\exists P.~v_2 = P(v_1)$).  Hence, $v_1$ is a
semantic ancestor of $v_2$.  Fig.~\ref{fig:syntactic-ancestor-2}
captures another way of establishing an ancestor relationship, namely by
merging branches. Version $v_{21}$ on branch $b_2$ is merged into
$v_{11}$ on $b_1$ to create $v_{12}$ on $b_1$. Versions $v_{11}$ and
\begin{wrapfigure}{l}{.4\textwidth}
\centering
\includegraphics[scale=0.35]{Figures/pushouts}
\caption{Commutative diagram representing the merge operation.}
\label{fig:pushouts}
\end{wrapfigure}
$v_{21}$ are now syntactic ancestors of $v_{12}$, but for them to be
semantic ancestors, the merge function has to enforce the relationship
explicitly. In other words, the result of merging  a pair of
semantic successors, $B$ and $C$, of an $A$, has to be an object $D$
that is a semantic successor of both $B$ and $C$. We now formalize
this intuition to define a mergeable type.
\begin{definition} [\bfseries Mergeable Type]
\label{def:mergeable-type}
A type $t$ is said to be mergeable, if and only if there exists a
function $M$ of type $t \times t \times t \rightarrow (t \rightarrow
t)\times(t \rightarrow t)$ that satisfies the following property:
$\forall (A, B, C\,:\,t).~ (\exists (P,Q\,:\, t \rightarrow t).~ B =
P(A) \conj C = Q(A)) \Rightarrow (\exists (P',Q': t\rightarrow
t).~M(A,B,C) = (P',Q') \conj  Q'(B) = P'(C))$
\end{definition}
Intuitively, a type $t$ is a mergeable type if, whenever there exist
two morphisms $P$ and $Q$ that map $A:t$ to $B:t$ and $C:t$, there
also exists morphisms $P'$ and $Q'$, that map both $B$ and $C$ to
$D:t$. This intuition is expressed visually in
Fig.~\ref{fig:pushouts}.  For a mergeable type $t$, we have the
guarantee that a sequence of values read by a thread from its branch
is sensible, as per the data type semantics:

@@

Rather than specifying merge ($M$) in terms of an LCA,
Definition~\ref{def:mergeable-type} specifies $M$ in terms of a pair
of morphisms.  This specification is useful to avoid the expression of certain
nonsensical merge functions, and thus enforce a useful branch-local
consistency property (Theorem~\ref{thm:branch-consistency}). For
instance, consider the counter data type from Sec.~\ref{sec:intro}. A
three-way merge function can be defined to return a constant
regardless of its arguments:
\begin{ocaml}
  let merge lca a b = 0
\end{ocaml}
Such a merge function allows the client to
witness the violation of the branch-local consistency property. For
example, the client may \C{sync} the counter, whose current local
value is $10$, and obtain $0$ as a result, thus witnessing a violation
of counter's monotonicity property. Fortunately,
Def.~\ref{def:mergeable-type} disallows this semantics. It is
impossible to define the function $M$ for the counter data type that
violates the monotonicity invariant, because no counter morphism
violates the invariant. In general, the specification of $M$
guarantees that the merge operation for a type $t$ preserves any
invariants (e.g., balancedness of a tree, sortedness of a list,
non-negativeness of an integer etc) that are preserved by all of $t$'s
morphisms.  The commutativity of the diagram in
Fig.~\ref{fig:pushouts} need not be enforced statically. It can
equally be verified at run-time by checking that the morphisms returned by $M$
map respective concurrent versions to the same value. The
operational semantics (Fig.~\ref{fig:opsem}) can be extended to this
effect.

@@

Merge functions for simple data types are easy to write. For instance,
the merge logic for counter is simple enough that it does not require
great effort to identify the $P$ and $Q$ that leda to the
concurrent counter versions, and transform them into $P'$ and $Q'$
such that the diagram in Fig.~\ref{fig:pushouts} commutes. For more
\begin{wrapfigure}{l}{.4\textwidth}
\centering
\includegraphics[scale=0.35]{Figures/pushouts-2}
\caption{Commutative diagram for Def.~\ref{def:mergeable-type-2}.}
\label{fig:pushouts-2}
\end{wrapfigure}
sophisticated data types, this process is more involved, and may require
the programmer to perform non-trivial reasoning. We now present an
alternative definition of a mergeable type that is stronger than
Def.~\ref{def:mergeable-type}, but also serves as a recipe to build
non-trivial merge functions.

@@

  \item There exists a function $M: (t \rightarrow t)^*\!\times\!(t
  \rightarrow t)^* \;\rightarrow\; (t \rightarrow t)\!\times\!(t
  \rightarrow t)$ that accepts a pair of minimal sequences of
  primitive morphisms, $P^*$ and $Q^*$, and returns a pair of morphisms,
  $P'$ and $Q'$, such that for any object $A:t$, $(Q' \circ
  \denot{P^*})(A) = (P' \circ \denot{Q^*})(A)$.  
\end{itemize}
The Commutativity diagram of Fig.~\ref{fig:pushouts-2} visualizes this
definition. 
\end{definition}

The definition asserts the existence of two functions for a mergeable
type $t$. First, the function $W$, which computes the \emph{edit
distance} between an object $A:t$ and its semantic successor $B:t$.
The sequence of primitive morphisms it returns ($P^*$) is called an
\emph{edit script}. Replaying the edit script on $A$ produces $B$
(i.e., $\denot{P^*}(A) = B$). Second, the definition asserts the
existence of a function $M$, which performs an \emph{operational
transformation} of the primitive morphisms in an edit script. An
operational transformation of a primitive morphism $P$ w.r.t a
morphism $Q$ is another primitive morphism $P'$ such that for any
object $A:t$, $P'$ has the same \emph{effect} on $Q(A)$ as $P$ has on
$A$. We will now consider examples of mergeable types that reify the
notions of edit scripts and operational transformations.
\begin{wrapfigure}{l}{.4\textwidth}
\centering
\includegraphics[scale=0.4]{Figures/string-eg}
\caption{Merging strings with edit scripts and operational
transformations}
\label{fig:string-eg}
\end{wrapfigure}
\paragraph{Strings} Strings are mergeable. Given a string ($s$) and
its semantic successor ($s'$), an edit script ($P^*$ or $Q^*$) is the
shortest sequence of \C{insert}, \C{delete}, and \C{substitute}
operations (a string's primitive morphisms), whose composition takes $s$
to $s'$. The length of the sequence is in fact the Levenshtein
distance~\cite{levenstein} between $s$ and $s'$, which can be computed
in polynomial time~\cite{wagner-fischer}.
A slight modification of Wagner-Fischer also lets us compute the edit
script. Fig.~\ref{fig:string-eg} illustrates.

>>>>

\begin{figure}
\centering
\includegraphics[scale=0.4]{Figures/pushouts}

\caption{Commutative diagram for Def.~\ref{def:mergeable-type}.}
\label{fig:pushouts}
\end{figure}

We now formally define what we mean by a mergeable type. The
definition serves as a guide to write merge functions for functional
data types, and elevate them to the level of replicated data types.

A functional data type library is as a tuple consisting of a type
($t$), and a (possibly infinite) collection of functions ($f$, $g$,
etc.) of type $t \rightarrow t$. We call the functions \emph{primitive
morphisms}\footnote{We borrow terminology from category theory for
convenience}. A morphism ($P$, $Q$, etc.) is either a primitive
morphism, or an associative composition of morphisms ($P \circ Q \circ
R$). An object $B:t$ is called a \emph{semantic successor} of $A:t$
(conversely, $A$ is called a \emph{semantic ancestor} of $B$) if and
only if there exists a morphism $P$ such that $P(A) = B$.

The aim of a three-way merge function over a type $t$ is to merge a
pair of semantic successors, $B:t$ and $C:t$, of an object $A:t$, into
another object $D:t$ such that the relationship between the semantic
successors and $D$ satisfies certain conditions. These conditions can
be understood by observing the \rulelabel{E-Pull-Wait} rule of of
Fig.~\ref{fig:opsem}, which applies the \C{merge} function to a pair
of concurrent versions ($v_1$ and $v_2$) and their least common
ancestor ($v$). Thus, the only relationship that exists between $v$
and $v_1$, and $v$ and $v_2$ is the syntactic ancestor relationship
that follows from the branching structure. The merge function, as
described above, however assumes that the concurrent versions are
semantic successors of their LCA. It is therefore essential to
maintain coherence between the syntactic and semantic ancestor
relations.

The ways in which syntactic ancestor relationships are created among
versions is captured in Fig.~\ref{fig:syntactic-ancestors}. Whenever an
object is pushed onto the branch, an ancestor relationship is created
between the previous version $v_1$, and the newly created version
$v_2$ (Fig.~\ref{fig:syntactic-ancestor-1}). However, since $v_2$ is
pushed by the thread after reading $v_1$ into the heap, it is
reasonable to assume that $v_2$ is a result of applying a morphism $P$
to $v_1$ (i.e., $\exists P.~v_2 = P(v_1)$).  Hence, $v_1$ is a
semantic ancestor of $v_2$.  Fig.~\ref{fig:syntactic-ancestor-2}
captures another way of establishing ancestor relationship, namely by
merging branches. Version $v_{21}$ on branch $b_2$ is merged into
$v_{11}$ on $b_1$ to create $v_{12}$ on $b_1$. Versions $v_{11}$ and
$v_{21}$ are now syntactic ancestors of $v_{12}$, but for them to be
semantic ancestors, merge function has to enforce the relationship
explicitly. In other words, the result of merging of a pair of
semantic successors, $B$ and $C$, of an $A$, has to be an object $D$
that is a semantic successor of both $B$ and $C$. We now formalize
this intuition to define a mergeable type.
\begin{definition} [\bfseries Mergeable Type]
\label{def:mergeable-type}
A type $t$ is said to be mergeable, if and only if there exists a
function $M$ of type $t \times t \times t \rightarrow (t \rightarrow
t)\times(t \rightarrow t)$ that satisfies the following property:
$\forall (A, B, C\,:\,t).~ (\exists (P,Q\,:\, t \rightarrow t).~ B =
P(A) \conj C = Q(A)) \Rightarrow (\exists (P',Q': t\rightarrow
t).~M(A,B,C) = (P',Q') \conj  Q'(B) = P'(C))$
\end{definition}
Intuitively, a type $t$ is a mergeable type if, whenever there exist
two morphisms $P$ and $Q$ that map $A:t$ to $B:t$ and $C:t$, there
must also exist morphisms $P'$ and $Q'$, the map both $B$ and $C$ to
$D:t$. This intuition is expressed visusally in
Fig.~\ref{fig:pushouts}.

For a mergeable type $t$, we have the guarantee that a sequence of
values read by a thread from its branch ``makes sense'' as per the
data type semantics:

@@

\begin{figure}
\centering
\includegraphics[scale=0.4]{Figures/pushouts-2}

\caption{Commutative diagram for Def.~\ref{def:mergeable-type-2}.}
\label{fig:pushouts-2}
\end{figure}

Def.~\ref{def:mergeable-type} specifies the merge function ($M$)
slightly differently; instead of returning a merged value like the
three-way merge function (\C{merge}), $M$ returns a pair of morphisms.
This specification is useful to preempt certain non-sensical merge
functions, and thus enforce the branch-local consistency property
(Theorem~\ref{thm:branch-consistency}). For instance, consider the
counter data type from Sec.~\ref{sec:intro}. A three-way merge
function can be defined to return a constant regardless of its
arguments:
\begin{ocaml}
  let merge lca a b = 0
\end{ocaml} Such a merge function allows the client (a thread) to
witness the violation of the branch-local consistency property. For
example, the client may \C{sync} the counter, whose current local
value is $10$, and obtain $0$ as a result, thus witnessing a violation
of counter's monotonicity property. Fortunately,
Def.~\ref{def:mergeable-type} disallows this semantics. It is
impossible to define the function $M$ for the counter data type that
violates the monotonicity invariant, because no counter morphism
violates the invariant. In general, the specification of $M$
guarantees that the merge operation for a type $t$ preserves any
invariants (e.g., balancedness of a tree, sortedness of a list,
non-negativeness of an integer etc) that are preserved by all of $t$'s
morphisms. On the hand, the commutativity of the diagram in
Fig.~\ref{fig:pushouts} may not be enforced statically. It can however
be verified at run-time by checking that the morphisms returned by $M$
map the respective concurrent versions to the same value. The
operational semantics (Fig.~\ref{fig:opsem}) can be extended to this
effect.

@@

Merge functions for simple data types are easy to write. For instance,
the merge logic for counter is simple enough that it doesn't require
explicit efforts to identify $P$ and $Q$ that led to the
concurrent counter versions, and transform them into $P'$ and $Q'$
such that the diagram in Fig.~\ref{fig:pushouts} commutes. For more
sophisticated data types, this process is more involved, and requires
the programmer to perform non-trivial reasoning. We now present an
alternative definition of a mergeable type that is stronger than
Def.~\ref{def:mergeable-type}, but also serves as a recipe to build
non-trivial merge functions.

@@

  \item There exists a function $M: (t \rightarrow t)^*\!\times\!(t
  \rightarrow t)^* \;\rightarrow\; (t \rightarrow t)\!\times\!(t
  \rightarrow t)$ that accepts a pair of minimal sequences of
  primitive morphisms, $P^*$ and $Q^*$, and returns a pair of morphisms,
  $P'$ and $Q'$, such that for any object $A:t$, $(Q' \circ
  \denot{P^*})(A) = (P' \circ \denot{Q^*})(A)$.  
\end{itemize}
Commutativity diagram of Fig.~\ref{fig:pushouts-2} visualizes this
definition. 
\end{definition}

The definition asserts the existence of two functions for a mergeable
type $t$. First, the function $W$, which computes the \emph{edit
distance} between an object $A:t$ and its semantic successor $B:t$.
The sequence of primitive morphisms it returns ($P^*$) is called an
\emph{edit script}. Replaying the edit script on $A$ produces $B$
(i.e., $\denot{P^*}(A) = B$). Second, the definition asserts the
existence of a function $M$, which performs an \emph{operational
transformation} of the primitive morphisms in an edit script. An
operational transformation of a primitive morphism $P$ w.r.t a
morphism $Q$ is another primitive morphism $P'$ such that for any
object $A:t$, $P'$ has the same \emph{effect} on $Q(A)$ as $P$ has on
$A$. We will now consider examples of mergeable types that reify the
notions of edit scripts and operational transformations.

\paragraph{Strings} Strings are mergeable. Given a string ($s$) and
its semantic successor ($s'$), an edit script ($P^*$ or $Q^*$) is the
shortest sequence of \C{insert}, \C{delete}, and \C{substitute}
operations (string's primitive morphisms), whose composition takes $s$
to $s'$. The length of the sequence is in fact the Levenshtein
distance~\cite{levenstein} between $s$ and $s'$, which can be computed
in polynomial time via Wagner-Fischer algorithm~\cite{wagner-fischer}.
A slight modification of Wagner-Fischer also lets us compute the edit
script. Fig.~\ref{fig:string-eg} illustrates.

### 2017-02-26 paper/model.tex
List eg.

The \drawsome example demonstrates the utility of mergeable data types
in building concurrent applications with conceptual sharing of
state. This example, however, implicitly makes certain assumptions
about the model and the underlying system, such as the existence of a
single least common ancestor (LCA) for any pair of versions, the
ability to access a previous version on any branch, and mergeability
of any two concurrent versions. Enforcing these guarantees in a fully
decentralized distributed setting requires addressing non-trivial
theoretical challenges described in the next section. Unlike the
canvas, the process of writing merge functions for data types with
non-trivial invariants could be quite involved. The following sections
describe a principled approach to deriving merge functions for any
data type given its invariants. Like \C{Canvas.merge}, merge functions
often need to compare multiple versions of a data type for structural
equality, which, when performed naively, incurs cost linear in the
size of versions. The representation of a versioned data type in \name
is optimized for performing comparisons that allows structural
equality to be determined in constant time. Such engineering details,
which contribute to the practicality of our programming model are also
discussed in subsequent sections.

>>>>

The \drawsome example demonstrates the utility of mergeable data types
and \name programming model in building concurrent applications with
conceptual sharing of state. The model lets concurrent computations be
composed around any mergeable type. As exemplified by \drawsome,
writing a three-way \C{merge} function is the only creative process in
lifting an OCaml data type to a mergeable type. We now present few
more examples of mergeable types, and isolate a recurring pattern in
their merge functions that serves as a guide to write merge functions
for more sophisticated types.

\begin{figure}
  \centering
  \includegraphics[scale=0.4]{Figures/list-eg}

  \caption{Lists of mergeable values are mergeable. }
  \label{fig:list-eg}
\end{figure}

\input{list-code}

{\bf Lists}. List, as an abstract data type, supports three
operations: \C{insert x i l}, that inserts an element \C{x} at
position \C{i} in the list \C{l}, \C{delete i l}, that deletes the
element at \C{i}'th position, and \C{subst i x l}, that substitutes
the element at \C{i}'th position with \C{x}. Lists of mergeable items
are mergeable. The signature of a mergeable list library is shown in
Fig.~\ref{fig:mlist-sig}. Note that the type of items in the list
(\C{A.t}) is also required to be mergeable. The signature
\C{MERGEABLE} will be discussed shortly.

The merge operation for lists is composed of two separate functions,
\C{edit\_seq} and \C{op\_transform}.
\begin{itemize}
  \item \C{edit\_seq} takes a pair of lists, \C{v} and $\C{v'}$, and
  computes the shortest sequence of list operations that need to be
  applied on \C{v} to obtain \C{v'}. Such a sequence is called an
  \emph{edit sequence}. The length of the sequence corresponds to the
  standard notion of the \emph{edit distance} between the two lists,
  which can be computed in polynomial time, for e.g., using the
  Wagner-Fisher algorithm~\cite{wagner-fischer}. A slight modification
  of Wagner-Fischer also lets us compute the edit sequence. The
  implementation in Fig.~\ref{fig:mlist} represents edits using the
  type \C{edit}. Constructors \C{I}, \C{D}, and \C{S} stand for
  \C{insert}, \C{delete}, and \C{subst}, respectively; \C{Nop} is
  there for technical reasons. The subst constructor also carries the
  \C{A.t} element that was substituted. Our implementation of
  Wagner-Fischer is more-or-less standard (hence, not shown), except
  that the edit sequence it returns orders \C{I}'s and \C{D}'s before
  \C{S}'s.  Fig.~\ref{fig:list-eg} illustrates edit sequences for a
  sample list. The sequence \C{[I(c,0); S(3,c,s)]} maps the list
  \C{[a;b;c]} to \C{[c;a;b;s]} because \C{subst 3 s (insert c 0
  [a;b;c]) = [c;a;b;s]}.
  \item  \C{op\_transform} takes a pair of edit sequences, $s_1$ and
  $s_2$, that map a list $v$ to two different lists, $v_1$ and $v_2$
  (e.g., Fig.~\ref{fig:list-eg}), and transforms $s_1$ to $s_1'$ such
  that $s_1'$ has the same effect on $v_2$ as $s_1$ had on $v$.  For
  instance, in Fig.~\ref{list-eg}, let $s_1$ be the edit sequence
  \C{[D(1); S(d,1)]}, which maps the list \C{[a;b;c]} ($v$) to
  \C{[a;d]} ($v_1$). The \C{D} edit deletes the first element of $v$.
  However, if $s_1$ is to be applied on the list \C{[c;a;b;s]}
  ($v_2$), which has a new element at position 0, the \C{D} edit must
  delete the the second element from $v_2$, if it were to have the
  same effect as it did on $v$. We say \C{D(1)} in $s_1$ has to be
  transformed w.r.t the sequence $s_2 = \C{[I(c,0);S(3,c,s)]}$.  The
  function \C{op\_transform} of Fig.~\ref{fig:mlist-xform} computes
  such \emph{operational transformation} for the sequence \C{mine}
  w.r.t the sequence \C{others}, and returns \C{mine'} as the
  transformed sequence.  
\end{itemize}

This example, however, implicitly makes certain assumptions
about the model and the underlying system, such as the existence of a
single least common ancestor (LCA) for any pair of versions, the
ability to access a previous version on any branch, and mergeability
of any two concurrent versions. Enforcing these guarantees in a fully
decentralized distributed setting requires addressing non-trivial
theoretical challenges described in the next section. Unlike the
canvas, the process of writing merge functions for data types with
non-trivial invariants could be quite involved. The following sections
describe a principled approach to deriving merge functions for any
data type given its invariants. Like \C{Canvas.merge}, merge functions
often need to compare multiple versions of a data type for structural
equality, which, when performed naively, incurs cost linear in the
size of versions. The representation of a versioned data type in \name
is optimized for performing comparisons that allows structural
equality to be determined in constant time. Such engineering details,
which contribute to the practicality of our programming model are also
discussed in subsequent sections.
